\bf Hierarchical Network Structure Promotes Dynamical Robustness

### 2015-05-20 tex/introduction.tex
update text

Here we determine the relationship between network hierarchy, a topological property, and the probability of \emph{robustness}, a dynamical property. Robustness is of interest in biological systems at all scales, and has been previously studied in the context of biochemical networks \cite{Alon1999,Shinar2010}, gene-regulatory networks \cite{VanNimwegen1999,Siegal2002,Draghi2010,Wagner2013}, and ecological networks \cite{Rohr2014}. Over physiological timescales, the robustness of a particular network state may be evaluated by determining its linear stability \cite{Davis1962}. Network states that are linearly stable, are robust to perturbations in the states of their components. For example, in the case of gene-regulatory networks, a state vector containing protein molecule counts or concentrations of each gene that is stable with respect to the dynamics of the network will exponentially suppress any relatively small modifications to the state of a gene and return to the initial stable state. Here we are interested in evaluating robustness over evolutionary as opposed to physiological timescales.

Over evolutionary timescales it is expected for there to be fluctuations, not only in the state, but in any aspect of the model specifying the dynamics of the network itself (i.e. parameters, structure of the rate functions, state variables, etc.). It is therefore also expected that network architectures managing to persist over such timescales may be required to do so with modifications to the location of their stable states. However, what must remain invariant on such evolutionary timescales is the higher-level constraint that the system possess a stable attractor at all associated to the basin selected at the initiation of its dynamics. This is necessary in order for particular lower-level networks to remain \emph{observable} long enough to serve as the substrate out of which relatively higher-level ones are constructed \cite{Simon2002}. What is important over at least moderate evolutionary timescales is then the conditional probability $R$, given that a system possesses a stable state somewhere in its state space, that upon modifications to its structure, it continues to possess such a stable state, regardless of where that state is located within the state space \ref{fig:robustnessconcept}. For the purpose of this investigation then, we quantify dynamical robustness as the probability of stability to potentially large perturbations of the model specifying the system dynamics for systems which have already been determined to be stable in the sense of linear stability analysis.

The maximally hierarchical network is considered to be the graph associated to the total ordering (\refsupp{} and \cite{Cormen2009}). An example of this for three system components is shown in \reffigscc{} top. We use a measure of hierarchy based upon the edit distance from this maximally hierarchical network \cite{Axenovich2011}. We demonstrate that systems with the most hierarchical network topology according to this definition exhibit maximal robustness and explain why this results from an invariance of robustness to particular kinds of transformations of the network topology. Moreover these results hold for networks of arbitrary size and independent of the probability distribution from which the strengths of interaction are sampled.
% \cite{WADDINGTON1942a,VanNimwegen1999,Siegal2002,Ciliberti2007b,Ciliberti2007,Wagner2013}

>>>>

Here we determine the relationship between network hierarchy, a topological property, and \emph{robustness}, a dynamical property. Robustness is of interest in biological systems at all scales, and has been previously studied in the context of biochemical networks \cite{Alon1999,Shinar2010}, gene-regulatory networks \cite{VanNimwegen1999,Siegal2002,Draghi2010,Wagner2013}, and ecological networks \cite{Rohr2014}. Over physiological timescales, the robustness of a particular network state may be evaluated by determining its linear stability \cite{Davis1962}. Network states that are linearly stable, are robust to perturbations in the states of their components. For example, in the case of gene-regulatory networks, a state vector containing protein molecule counts or concentrations of each gene that is stable with respect to the dynamics of the network will exponentially suppress any relatively small modifications to the state of a gene and return to the initial stable state. Here we are interested in evaluating robustness over evolutionary as opposed to physiological timescales.

Over evolutionary timescales it is expected for there to be fluctuations, not only in the state, but in any aspect of the model specifying the dynamics of the network itself (i.e. parameters, structure of the rate functions, etc.). It is therefore also expected that network architectures managing to persist over such timescales may be required to do so with modifications to the location of their stable states. However, what must remain invariant on such evolutionary timescales is the higher-level constraint that the system possess a relatively high overall probability of remaining in a stable state upon modifications to the underlying dynamical process. This is necessary in order for networks of lower-level components to exist, regardless of what state they exist in, long enough to serve as the substrate out of which networks of relatively higher-level components are constructed \cite{Simon2002}. What is important over at least moderate evolutionary timescales is then the conditional probability, $R$, given a system is in a stable state, that upon modifications to its structure in the context of environmental fluctuations, it remains stable, regardless of where the stable state is located within the state space \ref{fig:robustnessconcept}. For the purpose of this investigation then, we quantify dynamical robustness in this way.
% as the probability of stability to potentially large perturbations of the model specifying the system dynamics for systems which have already been determined to be linearly stable.

We demonstrate that systems exhibiting maximal robustness with respect to this definition have the most hierarchical network topology and explain why this results from an invariance of robustness to particular kinds of transformations of the network topology. For the purpose of formulating this result, the maximally hierarchical network is considered to be the graph associated to the total ordering (\refsupp{} and \cite{Cormen2009}). An example of this for three system components is shown in \reffigscc{} top. We use a measure of hierarchy based upon the edit distance from this maximally hierarchical network \cite{Axenovich2011}. Our results hold for networks of arbitrary size and are independent of the probability distribution from which the strengths of interaction are sampled.
% \cite{WADDINGTON1942a,VanNimwegen1999,Siegal2002,Ciliberti2007b,Ciliberti2007,Wagner2013}

### 2015-05-20 tex/methods.tex
update text

The rate equations for such a system are:
% \begin{widetext}
\begin{equation}\label{eq:crnarbitraryjacobian}
\begin{aligned}
\frac{dc_i}{dt} = &F_i = k_{\emptyset \to A_i} - k_{A_i \to \emptyset} c_i + k_{2A_i \to 3A_i} c_i^2 \\
&+ \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_j \to A_i + A_j} c_j - k_{A_i + A_j \to A_j} c_i c_j
\end{aligned}
\end{equation}
% \end{widetext}
The Jacobian at $\vec{x}^0$ is given as
% \begin{widetext}
\begin{align*}
\left. \frac{\partial F_i}{\partial c_i}\right|_{\vec{x}^0} &= - k_{A_i \to \emptyset} + 2 k_{2A_i \to 3A_i} - \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_i + A_j \rightarrow A_j}, & \\
\left. \frac{\partial F_i}{\partial c_j}\right|_{\vec{x}^0} &= k_{A_j \to A_i + A_j} - k_{A_i + A_j \to A_j}, &
\end{align*}
where $i \neq j$.
% \end{widetext}
By combining the equations $F_i(\vec{x}^0) = 0$ from \ref{eq:crnarbitraryjacobian} and $\frac{\partial F_i}{c_j}|_{\vec{x}^0} = M_{ij}$ we obtain the equivalent system of equations
\begin{align}
& k_{2A_i \to 3A_i} - k_{\emptyset \to A_i} = M_{ii} + \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_j \to A_i + A_j} \label{eq:jacobianconstraint1}  \\
& 2k_{2A_i \to 3A_i} - k_{A_i \to \emptyset} = M_{ii} + \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_i + A_j \to A_j} \label{eq:jacobianconstraint2}\\
& k_{A_j \to A_i + A_j} - k_{A_i + A_j \to A_j} = M_{ij} \label{eq:jacobianconstraint3}
\end{align}
We may solve these equations for the rate constants as follows.  We begin by solving \ref{eq:jacobianconstraint3} by either choosing $k_{A_i + A_j \to A_j} \ge 0$ and setting $k_{A_j \to A_i + A_j} = M_{ij} + k_{A_i + A_j \to A_j}$ when $M_{ij} \ge 0$ or choosing $k_{A_j \to A_i + A_j} \ge 0$ and setting $k_{A_i + A_j \to A_j} = k_{A_j \to A_i + A_j} - M_{ij}$ when $M_{ij} < 0$.  Pick
% \begin{widetext}
\begin{equation}
\begin{aligned}
k_{2A_i \to 3A_i} \ge \max \bigg(&0, M_{ii} + \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_j \to A_i + A_j}, \\
&M_{ii} + \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_i + A_j \to A_j} \bigg).
\end{aligned}
\end{equation}
% \end{widetext}
Then we may solve \ref{eq:jacobianconstraint1} for $k_{\emptyset \to A_i}$ and \ref{eq:jacobianconstraint2} for $k_{A_i \to \emptyset}$ and obtain non-negative answers. This demonstrates that arbitrary Jacobian matrices can arise from reaction network ensembles that allow for the possibility of at least those reactions in \ref{eq:arbitraryjacobianreactionnetwork}. Note that \ref{eq:crnarbitraryjacobian}, \ref{eq:jacobianconstraint1}, \ref{eq:jacobianconstraint2}, and \ref{eq:jacobianconstraint3} are linear in the parameter values. Therefore, any probability distribution on the elements of the Jacobian can be obtained from a probability distribution on the parameter values.

>>>>

The rate equations for such a system are:
% \begin{widetext}
\begin{equation}\label{eq:crnarbitraryjacobian}
\begin{aligned}
\frac{dc_i}{dt} = &F_i = k_{\emptyset \to A_i} - k_{A_i \to \emptyset} c_i + k_{2A_i \to 3A_i} c_i^2 \\
&+ \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_j \to A_i + A_j} c_j - k_{A_i + A_j \to A_j} c_i c_j
\end{aligned}
\end{equation}
% \end{widetext}
The Jacobian at $\vec{x}^0$ is given as
% \begin{widetext}
\begin{align*}
\left. \frac{\partial F_i}{\partial c_i}\right|_{\vec{x}^0} &= - k_{A_i \to \emptyset} + 2 k_{2A_i \to 3A_i} - \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_i + A_j \rightarrow A_j}, & \\
\left. \frac{\partial F_i}{\partial c_j}\right|_{\vec{x}^0} &= k_{A_j \to A_i + A_j} - k_{A_i + A_j \to A_j}, &
\end{align*}
where $i \neq j$.
% \end{widetext}
By combining the equations $F_i(\vec{x}^0) = 0$ from \ref{eq:crnarbitraryjacobian} and $\frac{\partial F_i}{\partial c_j}|_{\vec{x}^0} = M_{ij}$ we obtain the equivalent system of equations
\begin{align}
& k_{2A_i \to 3A_i} - k_{\emptyset \to A_i} = M_{ii} + \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_j \to A_i + A_j} \label{eq:jacobianconstraint1}  \\
& 2k_{2A_i \to 3A_i} - k_{A_i \to \emptyset} = M_{ii} + \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_i + A_j \to A_j} \label{eq:jacobianconstraint2}\\
& k_{A_j \to A_i + A_j} - k_{A_i + A_j \to A_j} = M_{ij} \label{eq:jacobianconstraint3}
\end{align}
We may solve these equations for the rate constants as follows.  We begin by solving \ref{eq:jacobianconstraint3} by either choosing $k_{A_i + A_j \to A_j} \ge 0$ and setting $k_{A_j \to A_i + A_j} = M_{ij} + k_{A_i + A_j \to A_j}$ when $M_{ij} \ge 0$ or choosing $k_{A_j \to A_i + A_j} \ge 0$ and setting $k_{A_i + A_j \to A_j} = k_{A_j \to A_i + A_j} - M_{ij}$ when $M_{ij} < 0$.  Pick
% \begin{widetext}
\begin{equation}
\begin{aligned}
k_{2A_i \to 3A_i} \ge \max \bigg(&0, M_{ii} + \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_j \to A_i + A_j}, \\
&M_{ii} + \sum_{\substack{1 \le j \le N \\ j \neq i \\ G_{ij} = 1}} k_{A_i + A_j \to A_j} \bigg).
\end{aligned}
\end{equation}
% \end{widetext}
Then we may solve \ref{eq:jacobianconstraint1} for $k_{\emptyset \to A_i}$ and \ref{eq:jacobianconstraint2} for $k_{A_i \to \emptyset}$ and obtain non-negative answers. This demonstrates that arbitrary Jacobian matrices can arise from reaction network ensembles that allow for the possibility of at least those reactions in \ref{eq:arbitraryjacobianreactionnetwork}. Note that \ref{eq:crnarbitraryjacobian}, \ref{eq:jacobianconstraint1}, \ref{eq:jacobianconstraint2}, and \ref{eq:jacobianconstraint3} are linear in the parameter values. Therefore, any probability distribution on the elements of the Jacobian can be obtained from a probability distribution on the parameter values.

### 2015-05-20 tex/overview.tex
update text

To determine stability, we linearize the equations of motion (\refsupp{}) about the
fixed point $\vec{x}^0$:
\begin{equation}\label{eq:lineardynsys}
\frac{d\vec{y}(t)}{dt} = A \vec{y}(t),
\end{equation}
where $\vec{y} = \vec{x} - \vec{x}^0$ and the $n \times n$ Jacobian matrix $A$ has components
$$
a_{ij} = \left. \frac{\partial F_i}{\partial x_j} \right|_{\vec{x} = \vec{x}^0}.
$$
The system defined by $F_i$ and $\vec{x}^0$ is dynamically stable if the eigenvalues of $A$ all have real parts less than zero and $A$ is then referred to as a stable matrix (\refsupp{}). The Lotka-Volterra model has
 \begin{equation}\label{eq:lotkavolterrajacobian}
   a_{ij} = \left\{
     \begin{array}{lr}
       r_i + b_{ii} n_i + \sum_{k=1}^{N} b_{ik} n_{k}, & i \neq j\\
       b_{ii} n_i & i=j
     \end{array}.
   \right.
\end{equation}

@@

In the process of biological evolution, the parameter values $\vec{p}$ or the form of the functions $F_i$ corresponding to all different types of networks considered in \ref{fig:biomodelexamples} are subject to, potentially drastic, modifications due, for example, to mutations or environmental fluctuations. These changes may alter the capacity of a given system to achieve a stable steady state. In order to evaluate this we consider a stochastic process whereby for a given dynamical system associated to a stable matrix $A$, we consider a perturbation of the system leading to $A'$ where $\adj(G_A) = \adj(G_{A'})$. This corresponds to a dynamical system whose connectivity is preserved, but where the underlying model is defined in terms of a different vector of parameters ${\vec{p}}\,'$ (\refsupp{}). We then ask, for each distinct network topology on $n$ components, what is the probability, given $A$ is a stable matrix, that $A'$ is also a stable matrix. This probability is what we refer to as dynamical robustness.

@@

>>>>

To determine stability, we linearize the equations of motion (\refsupp{}) about the
fixed point $\vec{x}^0$:
\begin{equation}\label{eq:lineardynsys}
\frac{d\vec{y}(t)}{dt} = A \vec{y}(t),
\end{equation}
where $\vec{y} = \vec{x} - \vec{x}^0$ and the $n \times n$ Jacobian matrix $A$ has components
$$
a_{ij} = \left. \frac{\partial F_i}{\partial x_j} \right|_{\vec{x} = \vec{x}^0}.
$$
The system defined by $F_i$ and $\vec{x}^0$ is dynamically stable if the eigenvalues of $A$ all have real parts less than zero and $A$ is then referred to as a stable matrix (\refsupp{}). This criterion can be checked equivalently in terms of conditions on the coefficients of the characteristic polynomials $\chi(A)$ associated to the systems described by matrices $A$ \cite{Gantmacher1959}. As an example of a Jacobian matrix, the Lotka-Volterra model has
 \begin{equation}\label{eq:lotkavolterrajacobian}
   a_{ij} = \left\{
     \begin{array}{lr}
       r_i + b_{ii} n_i + \sum_{k=1}^{N} b_{ik} n_{k}, & i \neq j\\
       b_{ii} n_i & i=j
     \end{array}.
   \right.
\end{equation}

@@

In the process of biological evolution, the parameter values $\vec{p}$ or the form of the functions $F_i$ corresponding to all different types of networks considered in \ref{fig:biomodelexamples} are subject to, potentially drastic, modifications due, for example, to mutations or environmental fluctuations. These changes may alter the capacity of a given system to achieve a stable steady state. In order to evaluate this we consider a stochastic process whereby for a given dynamical system associated to a stable matrix $A$, we consider a perturbation of the system leading to $A'$ where $\adj(G_A) = \adj(G_{A'})$ \ref{fig:robustnessconcept}. This corresponds to a dynamical system whose connectivity is preserved, but where the underlying model is defined in terms of a different vector of parameters ${\vec{p}}\,'$ (\refsupp{}). We then ask, for each distinct network topology on $n$ components, what is the probability, given $A$ is a stable matrix, that $A'$ is also a stable matrix. This probability is what we refer to as dynamical robustness.

@@

This argument also works when we resample more than one entry,
although the notation becomes more complicated.  Suppose that we
resample $m$ nodes.  Define
\begin{widetext}
\begin{equation*}
M = \left\{(m_0, m_1, \ldots, m_n) \,\bigg|\,
m = \sum_{i=0}^n m_i \quad\&\quad
m_0 \le \sum_{i=1}^{n-1} \sum_{j=i+1}^n \ell_{ij} \quad\&\quad
(\forall i \in \{1, \ldots, n\}) \; m_i < \connectivity_i \right\}.
\end{equation*}
\end{widetext}
Then, given $(m_0, m_1, \ldots, m_n) \in M$, there are ${m \choose
m_0, m_1, \ldots. m_n}$ ways of choosing $m_i$ links from $C_i$ and
$m_0$ links between strongly connected components.  Hence, our
weighted average becomes
\begin{widetext}
\begin{equation}
\langle R(\hbox{stab} (G), \mu') \rangle_m =
\frac{\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n}
      \left(m_0 + \sum\limits_{i=1}^n m_i \langle R(\hbox{stab} (C_i), \mu') \rangle_{m_i} \right)}
     {\sum\limits_{(m_0, m_1, \ldots m_n) \in M}
      {m \choose m_0, m_1, \ldots. m_n} m}
\end{equation}
\end{widetext}

As before, since $\langle R(\hbox{stab} (C_i), \mu') \rangle_{m_i} \le 1$, we may
increase $\langle R(\hbox{stab} (G), \mu') \rangle_m$ by increasing the maximum
possible value of $m_0$ while keeping the strongly connected
components the same.  Again, if we fix $\hier(G)$, the maximum
possible value of $m_0$ is $\sum_{(i,j) \in \hier(G)} v_i v_j$
whereas, if we allow it to vary, the maximum is $\frac{1}{2} ((\sum_{i=1}^n
v_i)^2 - \sum_{i=1}^n v_i^2)$, which is attained when $\hier(G_{max}) =
G_{tot}$.  Hence, we conclude that $\langle R(\hbox{stab} (G), \mu') \rangle_m \le
\langle R(\hbox{stab} (G_{\mathrm{max}}), \mu') \rangle_m$.
